---
title: "Additions"
author: "Bioinfo-Core @ MPI-IE"
date: "`r date() `"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    code_folding: hide
    toc: true
    toc_depth: 1
    toc_float: true
    df_print: paged
---

```{r load, message=FALSE}
library("DESeq2")
library("MASS")
```

## Example on disbalanced issues with colMeans trick

Let's have five WT samples:
2 WT = DMSO treated
3 WT = DRUG treated

and four KO samples:
2 KO = DMSO treated
2 KO = DRUG treated

and let's try to investigate the DRUG effect OVER the genotypes

```{r}
coldata <- data.frame(
    cond = c("DMSO", "DMSO", "DRUG", "DRUG", "DRUG", "DMSO", "DMSO", "DRUG", "DRUG"),
    genotype = c("WT", "WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO")
)
modmat <- model.matrix(~cond + genotype, coldata)

a <- modmat[coldata$cond == "DMSO", , drop = FALSE]
b <- modmat[coldata$cond == "DRUG", , drop = FALSE]

colMeans(b) - colMeans(a)
```

And compare this to what we get when we have BALANCED sampling:

```{r}
coldata <- data.frame(
    cond = c("DMSO", "DMSO", "DRUG", "DRUG", "DMSO", "DMSO", "DRUG", "DRUG"),
    genotype = c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO")
)
modmat <- model.matrix(~cond + genotype, coldata)

a <- modmat[coldata$cond == "DMSO", , drop = FALSE]
b <- modmat[coldata$cond == "DRUG", , drop = FALSE]

colMeans(b) - colMeans(a)
```


## Dispersion estimation under different conditions

First simulate data under Poisson

```{r}
df <- data.frame()
for (i in seq(1, 5000)) {
  # Sample a coef
  beta_cond = sample(c(rep(1, 11), 2, 4, 0.2, 0.5), 1)
  # sample a mean of counts
  mu = as.integer(runif(1, min=5, max=10000))
  df = rbind(
    df,
    c(rpois(4, lambda=mu), rpois(4, lambda=mu*beta_cond))
  )
}
colnames(df) <- c("WT_1", "WT_2", "WT_3", "WT_4", "KO_1", "KO_2", "KO_3", "KO_4")
coldata <- data.frame(
    genotype = factor(c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO"), levels=c("WT", "KO"))
)
dds <- DESeqDataSetFromMatrix(countData=df, colData=coldata, design=~genotype)
dds <- DESeq(dds)
plotDispEsts(dds)
```
Now simulate data under negative binomial

```{r}
df <- data.frame()
for (i in seq(1, 5000)) {
  # Sample a coef
  beta_cond = sample(c(rep(1, 11), 2, 4, 0.2, 0.5), 1)
  # sample a mean of counts
  mu = as.integer(runif(1, min=5, max=50000))
  df = rbind(
    df,
    c(rnegbin(4, mu = mu, theta=2000), rnegbin(4, mu = mu*beta_cond, theta=2000))
  )
}
colnames(df) <- c("WT_1", "WT_2", "WT_3", "WT_4", "KO_1", "KO_2", "KO_3", "KO_4")
coldata <- data.frame(
    genotype = factor(c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO"), levels=c("WT", "KO"))
)
dds <- DESeqDataSetFromMatrix(countData=df, colData=coldata, design=~genotype)
dds <- DESeq(dds)
plotDispEsts(dds)
```

