---
title: "Additions"
author: "Bioinfo-Core @ MPI-IE"
date: "`r date() `"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    code_folding: hide
    toc: true
    toc_depth: 1
    toc_float: true
    df_print: paged
---

```{r load, message=FALSE}
library("DESeq2")
library("ggplot2")
library("tidyverse")
library("ggExtra")
```

## Example on disbalanced issues with colMeans trick

Let's have five WT samples:  
2 WT = DMSO treated  
3 WT = DRUG treated  
  
and four KO samples:  
2 KO = DMSO treated  
2 KO = DRUG treated  

and let's try to investigate the DRUG effect OVER the genotypes.

```{r}
coldata <- data.frame(
    cond = c("DMSO", "DMSO", "DRUG", "DRUG", "DRUG", "DMSO", "DMSO", "DRUG", "DRUG"),
    genotype = c("WT", "WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO")
)
modmat <- model.matrix(~cond * genotype, coldata)

a <- modmat[coldata$cond == "DMSO", , drop = FALSE]
b <- modmat[coldata$cond == "DRUG", , drop = FALSE]

colMeans(b) - colMeans(a)
```

And compare this to what we get when we have BALANCED sampling:

```{r}
coldata <- data.frame(
    cond = c("DMSO", "DMSO", "DRUG", "DRUG", "DMSO", "DMSO", "DRUG", "DRUG"),
    genotype = c("WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO")
)
modmat <- model.matrix(~ cond * genotype, coldata)

a <- modmat[coldata$cond == "DMSO", , drop = FALSE]
b <- modmat[coldata$cond == "DRUG", , drop = FALSE]

colMeans(b) - colMeans(a)
```


## Dispersion examples

```{r load_data, message=FALSE, warning=FALSE}
data <- readr::read_tsv("data/myeloma/myeloma_counts.tsv")
data <- data %>% 
  column_to_rownames(var="gene_id") %>%
  mutate(across(where(is.double), as.integer))
```

#### Behavior of a true dataset

```{r plot gene sample means and variances, message=FALSE, warning=FALSE}
means <- data %>% select("BM_TG_1", "BM_TG_2", "BM_TG_3") %>% filter_all(any_vars(. != 0)) %>% apply(1, mean)
vars <- data %>% select("BM_TG_1", "BM_TG_2", "BM_TG_3") %>% filter_all(any_vars(. != 0)) %>% apply(1, var)
ggplot(data.frame(means,vars)) +
  geom_point(size=0.5, aes(x=means, y=vars)) +
  scale_y_log10(limits=c(1,1e10)) +
  scale_x_log10(limits=c(1,1e10)) +
  geom_abline(intercept = 0, slope = 1, color="red") +
  theme_minimal()

```

### Dispersion example when 'expected' distribution is followed

```{r disp-expected, warning=FALSE}
dds_normal <- makeExampleDESeqDataSet(
  n = 10000,
  m = 4,
  betaSD = 2,
  interceptMean = 4,
  interceptSD = 2
)
means <- counts(dds_normal)  %>% apply(1, mean)
vars <- counts(dds_normal) %>% apply(1, var)
ggplot(data.frame(means,vars)) +
  geom_point(size=0.5, aes(x=means, y=vars)) +
  scale_y_log10(limits=c(1,1e10)) +
  scale_x_log10(limits=c(1,1e10)) +
  geom_abline(intercept = 0, slope = 1, color="red") +
  theme_minimal()

```


```{r disp-expected plotest, message=FALSE}
dds_normal <- DESeq(dds_normal)
plotDispEsts(dds_normal)
```

```{r disp-expected ggplot, warning=FALSE}
p <- ggplot(data=mcols(dds_normal)) +
  geom_point(size=0.5, aes(x=baseMean, y=dispGeneEst)) +
  scale_y_log10() +
  scale_x_log10(limits=c(0.1,1e5)) +
  theme_minimal()
ggMarginal(p)
```
### Dispersion estimation when variance is very high

```{r disp-highvar, warning=FALSE}
dds_overdisp <- makeExampleDESeqDataSet(
  n = 10000,
  m = 4,
  betaSD = 2,
  interceptMean = 4,
  interceptSD = 2,
  dispMeanRel = function(x) x
)
means <- counts(dds_overdisp)  %>% apply(1, mean)
vars <- counts(dds_overdisp) %>% apply(1, var)
ggplot(data.frame(means,vars)) +
  geom_point(size=0.5, aes(x=means, y=vars)) +
  scale_y_log10(limits=c(1,1e10)) +
  scale_x_log10(limits=c(1,1e10)) +
  geom_abline(intercept = 0, slope = 1, color="red") +
  theme_minimal()
```

```{r, warning=FALSE}
dds_overdisp <- DESeq(dds_overdisp)
plotDispEsts(dds_overdisp)
```

```{r, warning=FALSE}
p <- ggplot(data=mcols(dds_overdisp)) +
  geom_point(size=0.5, aes(x=baseMean, y=dispGeneEst)) +
  scale_y_log10() +
  scale_x_log10(limits=c(0.1,1e5)) +
  theme_minimal()
ggMarginal(p)
```
```{r poisson, warning=FALSE}
dds_poisson <- makeExampleDESeqDataSet(
  n = 10000,
  m = 4,
  betaSD = 0,
  interceptMean = 4,
  interceptSD = 2,
  dispMeanRel = function(x) 0
)
means <- counts(dds_poisson)  %>% apply(1, mean)
vars <- counts(dds_poisson) %>% apply(1, var)
ggplot(data.frame(means,vars)) +
  geom_point(size=0.5, aes(x=means, y=vars)) +
  scale_y_log10(limits=c(1,1e10)) +
  scale_x_log10(limits=c(1,1e10)) +
  geom_abline(intercept = 0, slope = 1, color="red") +
  theme_minimal()
```

```{r, warning=FALSE}
dds_poisson <- DESeq(dds_poisson)
plotDispEsts(dds_poisson)
p <- ggplot(data=mcols(dds_poisson)) +
  geom_point(size=0.5, aes(x=baseMean, y=dispGeneEst)) +
  scale_y_log10() +
  scale_x_log10(limits=c(0.1,1e5)) +
  theme_minimal()
ggMarginal(p)
```


```{r, message=FALSE, warning=FALSE}
dds_normal2 <- makeExampleDESeqDataSet(
  n = 10000,
  m = 4,
  betaSD = 2,
  interceptMean = 8,
  interceptSD = 2
)

mixdf <- cbind(
  counts(dds_normal),
  counts(dds_normal2)
)
colnames(mixdf) <- c('wt_1', 'ko_1','wt_2', 'ko_2', 'wt_3', 'ko_3', 'wt_4' ,'ko_4')
metadf <- data.frame(
  c('wt','ko','wt','ko', 'wt','ko','wt','ko')
)
colnames(metadf) <- 'genotype'
rownames(metadf) <- c('wt_1', 'ko_1','wt_2', 'ko_2', 'wt_3', 'ko_3', 'wt_4' ,'ko_4')
metadf$genotype <- factor(metadf$genotype)
dds <- DESeqDataSetFromMatrix(mixdf, colData=metadf, design=~genotype)
means <- counts(dds)  %>% apply(1, mean)
vars <- counts(dds) %>% apply(1, var)
ggplot(data.frame(means,vars)) +
  geom_point(size=0.5, aes(x=means, y=vars)) +
  scale_y_log10(limits=c(1,1e10)) +
  scale_x_log10(limits=c(1,1e10)) +
  geom_abline(intercept = 0, slope = 1, color="red") +
  theme_minimal()
```
```{r, message=F, warning=F}
dds <- DESeq(dds)
plotDispEsts(dds)
```

```{r }
explode_var <- function(x){
  y = list()
  for (i in x){
    if (i > 10 & i < 1000){
      y <- c(y, 15)
    }
    else{
      y <- c(y, 4/i+0.1)
    }
  }
  return(unlist(y))
}
blowup_var <- function(x){
  y = list()
  for (i in x){
    if (i >= 50 & i <= 200){
      y <- c(y, 15)
    }
    else {
      y <- c(y, 0)
    }
  }
  return(unlist(y))
}

```



## INTERCEPTION

```{r read_data, message=FALSE, warning=FALSE}
# Get the data directory
dfile <- "data/myeloma/myeloma_counts.tsv"
data <- read_tsv(file=dfile)
data <- data %>%
  column_to_rownames("gene_id") %>%
  mutate(across(where(is.double), as.integer))

# Read samplesheet , detects conditions, generate the formula
mfile <- "data/myeloma/myeloma_meta.tsv"
metadata <- read_tsv(file=mfile)
metadata <- metadata %>% column_to_rownames("sample") 

metadata$condition <- as.factor(metadata$condition)
metadata$celltype  <- as.factor(metadata$celltype)
# sanity check
all(rownames(metadata) == colnames(data))
```


```{r}
dds <- DESeqDataSetFromMatrix(
  countData=data,
  colData=metadata,
  design= ~celltype + condition + celltype:condition
)
dds <- DESeq(dds)
resultsNames(dds)
```

```{r}
dds_noint <- DESeqDataSetFromMatrix(
  countData=data,
  colData=metadata,
  design= ~0 + celltype + condition + celltype:condition
)
dds_noint <- DESeq(dds_noint)
resultsNames(dds_noint)
```
```{r}
intres <- results(dds, contrast=c(0,1,0,0,0,0)) %>% data.frame() %>% drop_na()
nointres <- results(dds_noint, contrast=c(-1, 1, 0,0,0,0)) %>% data.frame() %>% drop_na()

ggplot() + 
  geom_point(aes(
    x=intres$log2FoldChange,
    y=nointres$log2FoldChange
  )
  ) + theme_minimal()

ggplot() + 
  geom_point(aes(
    x=intres$padj,
    y=nointres$padj
  )
  ) + theme_minimal()

```

