---
title: "deseq2 - applied"
author: "Bioinfo-Core @ MPI-IE"
date: "`r date() `"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
---

# Apply what was learned through the course.
  * Initial QC and visualisation.
  * Run the DESeq workflow
  * Interpret some of the results.

# Introduction
Dataset: MPPs

with:

> data/mpp/mpp_counts.tsv

> data/mpp/mpp_meta.tsv

Background:

We are looking in murine bone marrow cells. More specifically, the stem- and multipotent progenitor cells.
A (very expensive) experiment was set up, and there is RNA-seq data available three cellTypes:
 
 - HSCs
 - MPP1
 - MPP2

collected from both young and aged mice.
You and your lab are eager to see:

 - Did the experiment work ?
 - What are the effects of aging on the different cell populations ?

Hint: a list of cell-cycle genes (KEGG) are available under:

> data/genesets/KEGG_cellcycle.txt


<span style="color:red">Cave!: rownames in the count object are lowercase-capitalized, the KEGG_cellcycle ones are fully capitalized.</span> 

> Good luck and have fun!

***

# Hands-on
Load some libraries.

```{r add libraries, message=FALSE, warning=FALSE, paged.print=FALSE}
library("DESeq2")
library("dplyr")
library("tibble")
library("ggplot2")
library("pheatmap")
library("UpSetR")
library("tidyverse")
library("stringr")
```

Read in the data.
```{r read in data, message=FALSE, warning=FALSE, paged.print=FALSE}
dfile <- "data/mpp/mpp_counts.tsv"
data <- read_tsv(file=dfile)
data <- data %>% column_to_rownames("gene_id") 

mfile <- "data/mpp/mpp_meta.tsv"
metadata <- read_tsv(file=mfile)
metadata <- metadata %>% column_to_rownames("sample")
metadata$condition <- as.factor(metadata$condition)
metadata$celltype  <- as.factor(metadata$celltype)
```

Check our factors !
```{r factorLevels, message=FALSE, warning=FALSE, paged.print=FALSE}
print(metadata$celltype)
print(metadata$condition)
```
Relevel factors !
```{r relevel condition factor, message=FALSE, warning=FALSE, paged.print=FALSE}
levels(metadata$condition) <- c("Young","Aged")
print(metadata$condition)
```

## 1. Build DESeq object.

```{r build DESeq, message=FALSE, warning=FALSE, paged.print=FALSE}
dds <- DESeqDataSetFromMatrix(countData=data,
                              colData=metadata,
                              design = ~celltype*condition
                              )
```

## 2. plotPCA
```{r PCA, message=FALSE, warning=FALSE, paged.print=FALSE}
plotPCA(rlog(dds, blind=TRUE),
        intgroup=c("celltype", "condition")
        ) + theme_minimal()
```

## 3. run DESeq2

```{r Run deseq, message=FALSE, warning=FALSE, paged.print=FALSE}
dds <- DESeq(dds)
```

## 4. plot dispersion

```{r Dispersion, message=FALSE, warning=FALSE, paged.print=FALSE}
plotDispEsts(dds)
```

## 5. LRT for age.

```{r lrt-condition, message=FALSE, warning=FALSE}
# which model are we interested in as reduced?
dds_lrt <- DESeq(dds, test="LRT", reduced = ~celltype)
res_lrt <- results(dds_lrt)
```

## 6. results.
```{r parseResults, message=FALSE, warning=FALSE}
cleanDF <- function(RESobj){
  RESdf <- data.frame(RESobj) %>%
    drop_na() %>%
    arrange(padj)
  return(RESdf)
}

res_lrt_clean <- cleanDF(res_lrt)
head(res_lrt_clean)
```

## 7. Look at the 25 genes with lowest p-values
```{r heatmap_lrt, message=FALSE, warning=FALSE, fig.height=8, fig.weight=6}
normcounts <- counts(dds, normalized=TRUE) %>% data.frame()
top_25_genes <- res_lrt_clean %>% arrange(padj) %>% head(25) %>% rownames()
pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% top_25_genes),
  scale = 'row',
  main='Age effect - LRT top 25'
)

```

```{r heatmap_lrt_full, message=FALSE, warning=FALSE}
pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% rownames(res_lrt_clean %>% subset(padj < 0.05))),
  scale = 'row',
  main='Age effect - LRT',
  show_rownames = FALSE
)

```

```{r cell cycle, message=FALSE, warning=FALSE}
# Get cell cycle genes.
# Get the data directory
ccf <- "data/genesets/KEGG_cellcycle.txt"
cellcycle <- read_tsv(file=ccf, col_names=FALSE)
cellcycle <- cellcycle$X1

rownames(res_lrt_clean) <- res_lrt_clean %>% rownames() %>% toupper()
res_lrt_clean_cycle <- res_lrt_clean %>% subset(rownames(res_lrt_clean) %in% cellcycle) %>% subset(padj < 0.05)

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% str_to_title(rownames(res_lrt_clean_cycle))),
  scale = 'row',
  main='cell cycle genes - LRT',
  show_rownames = FALSE
)

```

```{r Age effect under cellTypes, message=FALSE, warning=FALSE}
# Remember our base levels and our coefficients.
print(resultsNames(dds))

# Get our pair-wise comparisons. Note we immediately clean up our results.
# Remember our matrix Trick!
mod_mat <- model.matrix(design(dds), colData(dds))

HSC_young <- colMeans(mod_mat[dds$celltype == 'HSC' & dds$condition == "Young",])
HSC_old <- colMeans(mod_mat[dds$celltype == 'HSC' & dds$condition == "Aged",])
MPP1_young <- colMeans(mod_mat[dds$celltype == 'MPP1' & dds$condition == "Young",])
MPP1_old <- colMeans(mod_mat[dds$celltype == 'MPP1' & dds$condition == "Aged",])
MPP2_young <- colMeans(mod_mat[dds$celltype == 'MPP2' & dds$condition == "Young",])
MPP2_old <- colMeans(mod_mat[dds$celltype == 'MPP2' & dds$condition == "Aged",])

# The easiest to fetch is the age effect in HSC cells.
ageEffect_HSC <- cleanDF( lfcShrink(dds, contrast = HSC_old-HSC_young , type="ashr") )

# Now let's get the age effect in MPP1 cells. (Which is the condition effect + our interaction term).
ageEffect_MPP1 <- cleanDF( lfcShrink(dds, contrast = MPP1_old-MPP1_young, type='ashr') )

# Lastly, let's get the age effect in MPP1 cells.
ageEffect_MPP2 <- cleanDF( lfcShrink(dds, contrast = MPP2_old-MPP2_young, type='ashr') )

```

```{r LTHSC, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% rownames(head(ageEffect_HSC, 25))) %>% dplyr::select(1,2,7,8),
  scale = 'row',
  main='age effect - HSCs',
  show_rownames = TRUE,
  annotation_row = ageEffect_MPP1[c(1:25), c('log2FoldChange'), drop=FALSE]
  )

# cell cycle
ageEffect_HSC_cycle <- ageEffect_HSC
rownames(ageEffect_HSC_cycle) <- ageEffect_HSC %>% rownames() %>% toupper()
ageEffect_HSC_cycle <- ageEffect_HSC_cycle %>% subset(rownames(ageEffect_HSC_cycle) %in% cellcycle) %>% subset(padj < 0.05)

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% str_to_title(rownames(ageEffect_HSC_cycle))) %>% dplyr::select(1,2,7,8),
  scale = 'row',
  main='cell cycle genes - HSC',
  show_rownames = TRUE
)

```

```{r MPP1, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% rownames(head(ageEffect_MPP1, 25))) %>% dplyr::select(3,4,9,10),
  scale = 'row',
  main='age effect - MPP1s',
  show_rownames = TRUE,
  annotation_row = ageEffect_MPP1[c(1:25), c('log2FoldChange'), drop=FALSE]
  )

# cell cycle
ageEffect_MPP1_cycle <- ageEffect_MPP1
rownames(ageEffect_MPP1_cycle) <- ageEffect_MPP1_cycle %>% rownames() %>% toupper()
ageEffect_MPP1_cycle <- ageEffect_MPP1_cycle %>% subset(rownames(ageEffect_MPP1_cycle) %in% cellcycle) %>% subset(padj < 0.05)

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% str_to_title(rownames(ageEffect_MPP1_cycle))) %>% dplyr::select(3,4,9,10),
  scale = 'row',
  main='cell cycle genes - MPP1s',
  show_rownames = TRUE
)
```

```{r MPP2, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% rownames(head(ageEffect_MPP2, 25))) %>% dplyr::select(5,6,11,12),
  scale = 'row',
  main='age effect - MPP2s',
  show_rownames = TRUE,
  annotation_row = ageEffect_MPP2[c(1:25), c('log2FoldChange'), drop=FALSE]
  )

# cell cycle
ageEffect_MPP2_cycle <- ageEffect_MPP2
rownames(ageEffect_MPP2_cycle) <- ageEffect_MPP2_cycle %>% rownames() %>% toupper()
ageEffect_MPP2_cycle <- ageEffect_MPP2_cycle %>% subset(rownames(ageEffect_MPP2_cycle) %in% cellcycle) %>% subset(padj < 0.05)

pheatmap(
  normcounts %>% subset(rownames(normcounts) %in% str_to_title(rownames(ageEffect_MPP2_cycle))) %>% dplyr::select(5,6,11,12),
  scale = 'row',
  main='cell cycle genes - MPP2s',
  show_rownames = TRUE
)
```