---
title: "DESeq2 Analysis with R: outline"
author: "Bioinfo-Core @ MPI-IE"
date: "`r date() `"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 1
    toc_float: true
    df_print: paged
---

This outline contains a template of a typical DESeq2 workflow - as discussed during the course.  
For specific projects, the parts on data exploration, quality controls and filtering are frequently more iterative and could require tweaking.


# Load Libraries
```{r load, message=FALSE}
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(ashr)

### define own convenience functions ###
my_volcano = function(res){
  # a low-key alternative to
  # EnhancedVolcano(res, lab = rownames(res), x = 'log2FoldChange', y = 'pvalue')
  res %>% 
    data.frame() %>%                                     # convert to data frame
    arrange(padj) %>%                                    # sort by padj 
    mutate(significant=padj < 0.05) %>%                  # define significant genes 
    mutate(rank=row_number()) %>%         # define rank as rownumber (after sorting)
    add_rownames("gene_name") %>%         # add rownames
    ggplot(aes(x=log2FoldChange, y=-log10(padj), color=significant)) + 
    geom_point() + 
    # label first 5 genes
    geom_text(aes(label=ifelse(rank<5,gene_name,'')), size = 3, hjust=0,vjust=-0.15) 
}
```


# Data, Metadata and Design
```{r import}
dfile <- "../data/mpp/mpp_counts.tsv"     
data <- read_tsv(dfile)
data <- data %>% column_to_rownames("gene_id") 

mfile <- "../data/mpp/mpp_meta.tsv"      
metadata <- read_tsv(mfile)
metadata <- metadata %>% column_to_rownames("sample") 
metadata$condition <- as.factor(metadata$condition)
metadata$celltype  <- as.factor(metadata$celltype)


my_design <- ~ celltype + condition + celltype:condition

all(rownames(metadata) == colnames(data))  # sanity check

dds <- DESeqDataSetFromMatrix(countData=data, colData=metadata, design= my_design)
```

# Data Exploration and Filters
```{r filtering}
# data exploration and QC: not shown
# ...

# remove genes
keep_genes <- rowSums(counts(dds)) > 1
dds <- dds[keep_genes,]

# process/filter samples as required
# ...

# rerun Data Exploration and QC
rld <- rlog(dds)
plotPCA(rld, intgroup=c("condition", "celltype"))

# keep transformed matrix A for later visualization
# scale rows (genes) for heatmap = transpose; (column) scale; transpose
A = assay(rld, normalized=TRUE) %>% t %>% scale %>% t 
```

# Run DESeq
```{r DEseq2}
dds <- DESeq(dds)                          

#Inspection
colData(dds)                                # sizefactors were added to metadata
plotDispEsts(dds, main="Dispersion plot")   # dispersion
# mcols(dds)                                # for the really curious: results of modeling
```

# Contrasts
Verify (base) levels and available coefficient (resultsNames).  
Define model matrix and contrasts systematically

```{r contrasts}
colData(dds)$condition
colData(dds)$celltype
resultsNames(dds)
mod_mat <- model.matrix(my_design, metadata)

# define contrast systematically
Aged_MPP2 <-  which(dds$condition=="Aged"  & dds$celltype=="MPP2")
Young_MPP2 <- which(dds$condition=="Young" & dds$celltype=="MPP2")
YvA_MPP2=colMeans(mod_mat[Young_MPP2,]) - colMeans(mod_mat[Aged_MPP2,]) 

YvA_MPP2
```

# Results
```{r results}
res <- lfcShrink(dds, contrast=YvA_MPP2, type = "ashr") 

# inspect results
res %>% head(5)                                        # first few results
summary(res)	                                         # summary for all genes
plotMA(res, ylim=c(-3,3))                              # MA-plot from DESeq2

plot(res$log2FoldChange, -log10(res$padj), cex=0.3)    # very simple volcano
#my_volcano(res)                                        # better?

# plot top-gene
plotCounts(dds, gene=which.min(res$padj), intgroup=c("condition","celltype"))  

# plot heatmap for top 15 genes
ntop <- 15
sorted_res <- res %>% data.frame() %>% arrange(padj)   # sort results by padj
top_genes <- sorted_res %>% head(ntop) %>% row.names() # top gene names
col_sel=c(Young_MPP2, Aged_MPP2)                       # select only samples in contrast
pheatmap(A[top_genes,col_sel], main="Top DE Genes (rlog)", annotation=metadata)

# write sorted results to file (add rowname as column)
fn="results.tsv"                        # might need a more informative filename
write_tsv(sorted_res %>% rownames_to_column("Gene"), file=fn)            
```

# SessionInfo
```{r sessionInfo}
sessionInfo()
```





